Xolotl: An Intuitive and Approachable Neuron \& Network Simulator in MATLAB

### 2018-07-11 paper/paper.tex
wrote new results section for mechanisms

In this section, we illustrate how "xolotl" can be used to generate a variety of models and simulate them. These examples have been chosen to illustrate various features of "xolotl", and are intended to serve as a template that researchers and educators can use for their own needs.

@@

In this section, we demonstrate how to simulate a Hodgkin-Huxley-like model, and how the tools built into our simulator make it easy to set up and integrate the model, and gain insight into the underlying biophysical mechanism. The schematic in Figure \ref{fig:figurehh} shows how the hierarchical structure of the ion channel populations within the soma is represented in the simulator, and how to create a compartment (here, labelled "HH") and add three populations of ion channels to it. These models of ion channels were obtained from \cite{liuModelNeuronActivitydependent1998} based on electrophysiological recordings from the lobster stomatogastric ganglion (\cite{turrigianoSelectiveRegulationCurrent1995}), and are part of the simulator. Adding an injected current and calling the built-in "plot" function plots the time series of membrane voltage. In the absence of injected current, the model is quiescent. When 0.2 nA is injected, the model tonically spikes (Figure \ref{fig:figurehh}A-B).  The \texttt{plot} function generates a voltage trace, which is colored by the dominant current (Figure \ref{fig:figurehh}A). Colors in the voltage trace indicate the strongest instantaneous inward current when the voltage is increasing, and strongest instantaneous outward current when the voltage is decreasing. These built-in features reveal that the dominant current during the upswing of every action potential is the sodium current, and the dominant current immediately after the peak of the action potential is the potassium current, but that the leak current contributes to depolarization following an action potential. This feature could be useful in quickly understanding the contributions of a number of ion channel types in a complex voltage trace from a more complicated neuron model.

@@

The model can be integrated for various amplitudes of injected current to determine its f-I curve, or the relationship between injected current and its firing rate (cite). Figure \ref{fig:figurehh}C shows the f-I curve of this model, obtained by repeated integration of the model.  Finally, the built-in \texttt{show} function can plot activation ($m$) and inactivation ($h$) curves and voltage-dependent timescales of any channel type in the simulator (Figure \ref{fig:figurehh}D-G). These plots reveal that activation kinetics of the "NaV" channels are much faster than that of the "Kd" channels (Figure \ref{fig:figurehh}F), which facilitates the transient depolarization in an action potential. In summary, the simulator allows the user to construct and integrate this model in a few lines of code, and provides a rich visualization of the dynamics of the model.

@@

Voltage clamping is a experimental technique where a amplifier is configured to inject the appropriate amount of current through a electrode to maintain the voltage of a cell at a desired value (cite D\&A). Under this paradigm, the membrane voltage is "clamped" or fixed to a desired value, permitting the study of voltage-dependent ion channels, since the sum of all currents through the population of ion channels in the cell is equal and opposite to the current injected by the clamp. By combining voltage clamp with the use of pharmacological agents to block all channels but the one of interest, the voltage-sensitivity of an ion channel population can be characterized. (insert all those citations as needed).

"xolotl" can reproduce such a voltage clamp experiment $in silico$. Figure \ref{fig:figureclamp} illustrates how a simple model with a single compartment and a single ion channel type can be set up and clamped to a desired voltage.  Integrating the model yields the current required to clamp the cell at that voltage. Here, we use a delayed-rectifier potassium conductance (\cite{liuModelNeuronActivitydependent1998}) and simulate a voltage-clamp experiment whose goal is to infer the activation function of this channel. First, the cell is clamped to a number of different voltages (Fig. \ref{fig:figureclamp}A) and the resultant clamping currents are measured by integrating the model (Fig. \ref{fig:figureclamp}B). The asymptotic clamping currents depend on the clamping voltage in a nonlinear manner (Fig. \ref{fig:figureclamp}C), since the open probabilities of the channel are functions of the membrane voltage. Assuming the reversal potential is known, Eq. \eqref{eq:2} can be used to solve for the total conductance of the channel as a function of the clamping voltage (Fig. \ref{fig:figureclamp}D). Finally, a sigmoid can be fit to the normalized conductance-voltage curves to obtain the activiation function of the ion channel population (Fig. \ref{fig:figureclamp}E-F). "xolotl" can therefore be used to describe graphically the theoretical underpinnings of ion channel characterization through voltage clamp and can serve as an effective pedagogical tool in computational neuroscience.

@@

"xolotl" can implement homeostatic tuning rules as integral control. The controller computes an error signal (typically a function of intracellular calcium concentration), and adjusts the conductance or synapse it controls accordingly (\cite{olearyCorrelationsIonChannel2013}). In "xolotl", integral controllers are "cpplab" objects added to the conductance or synapse they regulate.

In a demonstration adapted from \cite{olearyCorrelationsIonChannel2013}, integral control changes maximal conductances to bring a neuron from quiescence into a bursting regime. Calcium sensors supervene on maximal conductance density (Figure \ref{fig:figureintegralcontrol}) to change neuronal activity. Each conductance in the "xolotl" structure contains a calcium-sensitive controller (Figure \ref{fig:figureintegralcontrol}B-C). Maximal conductances increase from random initial conditions to a set which elicits the desired network output by minimizing the error signal (Figure \ref{fig:figureintegralcontrol}D-F).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		SNAPSHOTS
%
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

>>>>

In this section, we illustrate how "xolotl" can be used to generate a variety of models and simulate them. These examples have been chosen to illustrate various features of "xolotl", and are intended to serve as templates that researchers and educators can build upon.

@@

In this example, we demonstrate how to simulate the spiking activity of a Hodgkin-Huxley-like model, and how the tools built into our simulator make it easy to set up and integrate the model, and gain insight into the underlying biophysical mechanism. This simple model consists of a single-compartment neuron with three types of conductances  (Figure \ref{fig:figurehh}A). This heirarchical organization of the neuron is mirrored in the structure of the model: an object of type "compartment" is used to represent the cell body, and three objects of type "conductance" are used to model the three populations of ion channels (Figure \ref{fig:figurehh}B). These models of ion channels were obtained from \cite{liuModelNeuronActivitydependent1998} based on electrophysiological recordings from the lobster stomatogastric ganglion (\cite{turrigianoSelectiveRegulationCurrent1995}), and are part of the simulator. The syntax of constructing the neuron model (using "add" statements) preserves this herirarchical organization, and the code snippet in (Figure \ref{fig:figurehh}C) shows how this model can be created using just a few lines of code.

Adding an injected current and calling the built-in "plot" function plots the time series of membrane voltage. In the absence of injected current, the model is quiescent. When 0.2 nA is injected, the model tonically spikes (Figure \ref{fig:figurehh}A-B).  The \texttt{plot} function displays a voltage trace colored by the dominant current (Figure \ref{fig:figurehh}A). Colors in the voltage trace indicate the strongest instantaneous inward current when the voltage is increasing, and strongest instantaneous outward current when the voltage is decreasing. These built-in features reveal that the dominant current during the upswing of every action potential is the sodium current, and the dominant current immediately after the peak of the action potential is the potassium current, but that the leak current contributes to depolarization following an action potential. This feature could be useful in quickly understanding the contributions of a number of ion channel types in a complex voltage trace from a more complicated neuron model.

@@

The model can be integrated for various amplitudes of injected current to determine its f-I curve, or the relationship between injected current and its firing rate (cite). Figure \ref{fig:figurehh}E shows the f-I curve of this model, obtained by repeated integration of the model.  Finally, the built-in \texttt{show} function can plot activation ($m$) and inactivation ($h$) curves and voltage-dependent timescales of any channel type in the simulator (Figure \ref{fig:figurehh}D-G). These plots reveal that activation kinetics of the "NaV" channels are much faster than that of the "Kd" channels (Figure \ref{fig:figurehh}F), which facilitates the transient depolarization in an action potential. In summary, the simulator allows the user to construct and integrate this model in a few lines of code, and provides a rich visualization of the dynamics of the model.

@@

Voltage clamping is a experimental technique where a amplifier is configured to inject the appropriate amount of current through a electrode to maintain the voltage of a cell at a desired value (cite D\&A). Under this paradigm, the membrane voltage is "clamped" or fixed to a desired value, permitting the study of voltage-dependent ion channels, since the sum of all currents through the population of ion channels in the cell is equal and opposite to the current injected by the clamp (Figure \ref{fig:figureclamp}A). By combining voltage clamp with the use of pharmacological agents to block all channels but the one of interest, the voltage-sensitivity of an ion channel population can be characterized. (alec:insert all those citations as needed).

"xolotl" can reproduce such a voltage clamp experiment $in silico$. Figure \ref{fig:figureclamp}B illustrates how a simple model with a single compartment and a single ion channel type can be set up and clamped to a desired voltage.  Integrating the model yields the current required to clamp the cell at that voltage. Here, we use a delayed-rectifier potassium conductance (\cite{liuModelNeuronActivitydependent1998}) and simulate a voltage-clamp experiment whose goal is to infer the activation function of this channel. First, the cell is clamped to a number of different voltages (Fig. \ref{fig:figureclamp}C) and the resultant clamping currents are measured by integrating the model (Fig. \ref{fig:figureclamp}D). The asymptotic clamping currents depend on the clamping voltage in a nonlinear manner (Fig. \ref{fig:figureclamp}E), since the open probabilities of the channel are functions of the membrane voltage. Assuming the reversal potential is known, Eq. \eqref{eq:2} can be used to solve for the total conductance of the channel as a function of the clamping voltage (Fig. \ref{fig:figureclamp}F). Finally, a sigmoid can be fit to the normalized conductance-voltage curves to obtain the activiation function of the ion channel population (Fig. \ref{fig:figureclamp}G-H). "xolotl" can therefore be used to describe graphically the theoretical underpinnings of ion channel characterization through voltage clamp and can serve as an effective pedagogical tool in computational neuroscience.

@@

So far, the models we described only considered the voltage dynamics of a cell (the solution to Eq. \ref{eq:1}). However, real neurons possess a several dynamical features, stemming from a number of intracellular mechanisms. "xolotl" makes it possible to model and include arbitrary intracellular mechanisms. In our simulator, these intracellular mechanisms are represented by the "mechanism" object, and can be bound to "compartments", "conductances", and other object types.

One of the most important intracellular mechanisms is the cytosolic buffering of Calcium and its influx through voltage-gated Calcium channels (VGCCs). Figure \ref{fig:figuremechanism}A shows a model of a single-compartment model with 8 populations of ion channels (alec: cite Liu). Without any mechanism for Calcium influx or buffering, the intracellular Calcium levels in this model do not change (Figure \ref{fig:figuremechanism}B) and the model tonically spikes (Figure \ref{fig:figuremechanism}C). A common method of modelling Calcium buffering and influx is to use an differential equation that increases intracellular Calcium with the current through Calcium channels, and relaxes back to a baseline value (alec: cite Liu, Prinz, DayanAbbott) (Figure \ref{fig:figuremechanism}D). This mechanism exists in the "xolotl" library as "CalciumMech1" and can be added to the model using a simple "add" statement (Figure \ref{fig:figuremechanism}E). The intracellular Calcium in the model now oscillates periodically (Figure \ref{fig:figuremechanism}F), synchronised to bursts in action potentials in this cell (Figure \ref{fig:figuremechanism}G).

Another intracellular mechanism of interest is a neuron's ability to regulate its electrical activity by controlling the spectrum of ion channels its expresses. Here, we will show "xolotl" can be used to model a recently proposed model of a homeostatic feedback system that controls the transcription rates of ion channels with the integral of an error signal derived from the intracellular Calcium concentration and some target Calcium (alec: cite Tim) (Figure \ref{fig:figuremechanism}H). Since this mechanism affects each ion channel population individually, object corresponding to this mechanism is added to each "condutance" object in the neuron (Figure \ref{fig:figuremechanism}I). Setting all maximal conductance densities to some low value and integrating the model shows that the intracellular Calcium levels rise over time and approach the target (Figure \ref{fig:figuremechanism}J), while all conductance densities increase and then remain bounded (Figure \ref{fig:figuremechanism}K). Examining the voltage dynamics of the cell reveal that it transitions from silence to truncated bursts of action potentials to periodic bursting as this mechanism regualtes the neuron's ion channel spectrum.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%
%
%		SNAPSHOTS
%
%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
